---
title: "summitSNOTEL"
author: "Gordon Gianniny"
date: "2026-08-09"
output: html_document
---

```{r setup, include=F, echo = F}
library(tidyverse)
library(broom)
library(ggplot2)
library(lubridate)
library(RColorBrewer)
library(here)
library(patchwork)
library(snotelr)
library(zoo)
library(dataRetrieval)
library(cpop)
```

# Overview

This script reads in data from SNOTEL stations in Summit County, UT, reformats the data, and draws plots showing: 

  A. Change in snowpack over time
  
  B. Ratio of snow to total precip over time
  
Sections: 

  1. Read in SNOTEL Data
  2. Data re-formatting & cleanup
  3. Snowpack change over time plots
  4. Snow:total precip ratio
    + Calculate snow:total precip ratio
    + Draw plots
    
All plots are stored in the "outputs>plots" folder within the project directory. 

##1. Read in SNOTEL Data

Import data from Snotel using the snotelr package:

Get metadata for sites in Summit County:

```{r}
snotelMd <- snotel_info(network="sntl")%>%
  filter(county == "Summit"&state == "UT") #Info for sites in Summit county UT

siteIds <- c(snotelMd$site_id)
```

Data for all sites in Summit County:

```{r}
snotelData <- snotel_download(site_id = siteIds, network = "sntl", internal = T)%>%
  mutate(date = ymd(date), site_name = as.factor(site_name))
```

##2. Data re-formatting & cleanup

Dropping extra columns:

```{r}
snotelClean <- snotelData %>%
  select(!c(network, state, description, latitude, longitude, county, site_id))
```

##3. Snowpack change over time plots

###A. Calculate relevant snowpack variables for each year: 

```{r}
snotelStats <- snotelClean%>%
  mutate(year = year(date), doy = yday(date), month = month(date))%>%
  mutate(waterYear = ifelse(month<10, year, year+1))%>%
  group_by(site_name, waterYear)%>%
  summarize(sweMax = max(snow_water_equivalent, na.rm = T), 
            sweMax_doy = doy[which.max(snow_water_equivalent)])
```

##4. Snow:total precip ratio

### A. Calculate snow:total precip ratio

### B. Draw plots

