---
title: "summitSNOTEL"
author: "Gordon Gianniny"
date: "2026-08-09"
output: html_document
---

```{r setup, include=F, echo = F}
library(tidyverse)
library(broom)
library(ggplot2)
library(lubridate)
library(RColorBrewer)
library(here)
library(patchwork)
library(snotelr)
library(zoo)
library(dataRetrieval)
library(cpop)
```

# Overview

This script reads in data from SNOTEL stations in Summit County, UT, reformats the data, and draws plots showing: 

  A. Change in snowpack over time
  
  B. Ratio of snow to total precip over time
  
Sections: 

  1. Read in SNOTEL Data
  2. Data re-formatting & cleanup
  3. Snowpack change over time plots
  4. Snow:total precip ratio
    + Calculate snow:total precip ratio
    + Draw plots
    
All plots are stored in the "outputs>plots" folder within the project directory. 

##1. Read in SNOTEL Data

Import data from Snotel using the snotelr package:

Get metadata for sites in Summit County:

```{r}
snotelMd <- snotel_info(network="sntl")%>%
  filter(county == "Summit"&state == "UT") #Info for sites in Summit county UT

siteIds <- c(snotelMd$site_id)
```

Data for all sites in Summit County:

```{r}
snotelData <- snotel_download(site_id = siteIds, network = "sntl", internal = T)%>%
  mutate(date = ymd(date), site_name = as.factor(site_name))
```

##2. Data re-formatting & cleanup

Dropping extra columns & incomplete years; adding water year and DOY:

```{r}
snotelClean <- snotelData %>%
  select(!c(network, state, description, latitude, longitude, county, site_id))%>% #drop extra cols
  mutate(year = year(date), doy = yday(date), month = month(date))%>% #add year, day of year, and month columns
  group_by(site_name, year)%>% #group by site and year
  mutate(doyMin = min(doy), 
         doyMax = max(doy))%>% #calculate max and min day of year for each site/year
  ungroup()%>%
  filter(!doyMax<365)%>% #drop incomplete years
  mutate(waterYear = ifelse(month<10, year, year+1))%>% #Calculate water year
  mutate(waterDOY = ifelse(doy>=doyMax-91, 
                           doy-(doyMax-92), 
                           doy+92)
         )%>% # #use min/max DOY to calculate water day of year
  mutate(siteElev = paste("(", elev, " m", ")", " - ", site_name, sep =""))

```

##3. Snowpack change over time plots

###A. Calculate relevant snowpack variables for each year: 

```{r}
snotelStats <- snotelClean%>%
  filter(!doyMin>1, !doyMax<182)%>% #drop years that start after jan 1 or end before jul 1
  filter(doy>1&doy<182)%>% #extract Jan 1 to July 1
  group_by(site_name, elev, year, siteElev)%>% #group by site and year
  summarize(sweMax = max(snow_water_equivalent, na.rm = T), 
            sweMax_doy = doy[which.max(snow_water_equivalent)], 
            sweZero = yday(date[which.min(snow_water_equivalent)][1]))%>% #calculate max swe value, max swe Day of year, 1st day with zero swe
  mutate(meltDays = sweZero-sweMax_doy)%>% #calculate number days between peak swe and max swe
  pivot_longer(!c(site_name, elev, year, siteElev), names_to = "metricName", values_to = "metricValue") #pivot longer for plotting
```

###B. Site-by-site - Nest data and run basic LM's

```{r}
snotelNest <- snotelStats %>%
  nest(data = -c(site_name, elev, siteElev, metricName)) #create new nested dataframe with columns for site and metricName, one column for data - data entries for each site include year and metricValues. 
```

Run basic models for each site and metric of interest: 

```{r}
sModels <- snotelNest %>%
  mutate(model = map(data, ~lm(metricValue~year, data = .)%>% #run a model of temp~year for each site (referenced by "."); store results in new column called "model"
                     tidy))%>% #tidy function to construct a summary table with model results (seperate tables for each site)
  unnest(model) %>% #unnest model column to get rows for each site, columns for model metrics
  filter(term == "year")  #remove intercept term to just get info on year for each site
sModels
```

Add P-values for plotting: 

```{r}
swePV <- merge(snotelStats, 
                     select(sModels, site_name, metricName, p.value)
                     )%>%
  mutate(metricPrint = case_when(
    metricName == "sweMax" ~ "Maximum Recorded SWE, mm", 
    metricName == "sweMaxDOY" ~"Day of year of max. recorded SWE", 
    metricName == "sweZero" ~"First day of year w/ zero recorded SWE",
    metricName == "meltDays" ~"Melt duration, days"
  ))%>%
  mutate(siteMetric = paste(site_name, metricPrint, sep = " - "))

sPVs <- swePV%>%
  group_by(site_name, metricName, elev, siteElev)%>%
  summarize(x = mean(year), label = round(unique(p.value), 2))
```

###C. Draw and export site-by-site plots

Prepping color pallate:

```{r}
pal1 <- brewer.pal(11,"BrBG")

elevPal <- c(pal1[1:5], pal1[7:11])
```

Maximum SWE plot:

```{r}
swe.max.plot <- ggplot(filter(snotelStats, metricName == "sweMax"), aes(x = year, y = metricValue, color = elev))+
  geom_point()+
  geom_smooth(se = T, method = "lm", linetype = "dashed")+
  theme_classic()+
  labs(x = "Water Year", y = "Maximum recorded SWE, mm", color = "Elevation, m")+
  scale_color_continuous(palette = elevPal)+
  geom_text(data = filter(sPVs, metricName == "sweMax"), aes(x = x, label = paste("P=", label)), y = Inf, vjust = 2)+
  facet_wrap(~siteElev, scales = "free", labeller = label_wrap_gen(width = 20))+
  theme(axis.text = element_text(size = 11), 
        axis.title = element_text(size = 11, face = "bold"), 
        strip.text = element_text(size = 10, face = "bold"), 
        legend.position = "bottom")


swe.max.plot

#Save plot:
ggsave("SWE_max.jpg", swe.max.plot, device = "jpeg", path = here("outputs", "plots"), units = "in", dpi = "retina", height = 8, width = 11)
```

Maximum SWE DOY plot:

```{r}
swe.doy.plot <- ggplot(filter(snotelStats, metricName == "sweMax_doy"), aes(x = year, y = metricValue, color = elev))+
  geom_point()+
  geom_smooth(se = T, method = "lm", linetype = "dashed")+
  theme_classic()+
  labs(x = "Water Year", y = "Maximum recorded SWE day of year", color = "Elevation, m")+
  scale_color_continuous(palette = elevPal)+
  geom_text(data = filter(sPVs, metricName == "sweMax_doy"), aes(x = x, label = paste("P=", label)), y = Inf, vjust = 2)+
  facet_wrap(~siteElev, scales = "free", labeller = label_wrap_gen(width = 20))+
  theme(axis.text = element_text(size = 11), 
        axis.title = element_text(size = 11, face = "bold"), 
        strip.text = element_text(size = 10, face = "bold"), 
        legend.position = "bottom")


swe.doy.plot

#Save plot:
ggsave("SWE_max_doy.jpg", swe.doy.plot, device = "jpeg", path = here("outputs", "plots"), units = "in", dpi = "retina", height = 8, width = 11)
```

Melt days plot:

```{r}
melt.plot <- ggplot(filter(snotelStats, metricName == "meltDays"), aes(x = year, y = metricValue, color = elev))+
  geom_point()+
  geom_smooth(se = T, method = "lm", linetype = "dashed")+
  theme_classic()+
  labs(x = "Water Year", y = "# days from peak SWE to zero SWE", color = "Elevation, m")+
  scale_color_continuous(palette = elevPal)+
  geom_text(data = filter(sPVs, metricName == "meltDays"), aes(x = x, label = paste("P=", label)), y = Inf, vjust = 2)+
  facet_wrap(~siteElev, scales = "free", labeller = label_wrap_gen(width = 20))+
  theme(axis.text = element_text(size = 11), 
        axis.title = element_text(size = 11, face = "bold"), 
        strip.text = element_text(size = 10, face = "bold"), 
        legend.position = "bottom")


melt.plot

#Save plot:
ggsave("melt_days.jpg", melt.plot, device = "jpeg", path = here("outputs", "plots"), units = "in", dpi = "retina", height = 8, width = 11)
```

### D. Overall/combined plot: 

```{r}
combNest <- snotelStats %>%
  nest(data = -c(metricName)) #create new nested dataframe with columns for site and metricName, one column for data - data entries for each site include year and metricValues. 
```

Run basic models for each site and metric of interest: 

```{r}
combModels <- combNest %>%
  mutate(model = map(data, ~lm(metricValue~year, data = .)%>% #run a model of temp~year for each site (referenced by "."); store results in new column called "model"
                     tidy))%>% #tidy function to construct a summary table with model results (seperate tables for each site)
  unnest(model) %>% #unnest model column to get rows for each site, columns for model metrics
  filter(term == "year")  #remove intercept term to just get info on year for each site
combModels
```

Add P-values for plotting: 

```{r}
combPV <- merge(snotelStats, 
                     select(combModels, metricName, p.value)
                     )%>%
  mutate(metricPrint = case_when(
    metricName == "sweMax" ~ "Maximum Recorded SWE, mm", 
    metricName == "sweMaxDOY" ~"Day of year of max. recorded SWE", 
    metricName == "sweZero" ~"First day of year w/ zero recorded SWE",
    metricName == "meltDays" ~"Melt duration, days"
  ))

cPVs <- combPV%>%
  group_by(metricName)%>%
  summarize(x = mean(year), label = round(unique(p.value), 6))
```

#### i. Max SWE Plot:

```{r}
swe.max.comb <- ggplot(filter(snotelStats, metricName == "sweMax"), aes(x = year, y = metricValue, color = elev))+
  geom_point()+
  geom_smooth(se = T, method = "lm", linetype = "dashed", color = "black")+
  theme_classic()+
  labs(x = "Water Year", y = "Maximum recorded SWE, mm", color = "Site elevation, m")+
  scale_color_continuous(palette = elevPal)+
  geom_text(data = filter(cPVs, metricName == "sweMax"), aes(x = x, label = paste("P=", label)), y = Inf, vjust = 2, color = "black")+
  theme(axis.text = element_text(size = 11), 
        axis.title = element_text(size = 11, face = "bold"), 
        strip.text = element_text(size = 10, face = "bold"), 
        legend.position = "bottom")


swe.max.comb

#Save plot:
ggsave("SWE_max_comb.jpg", swe.max.comb, device = "jpeg", path = here("outputs", "plots"), units = "in", dpi = "retina", height = 8, width = 11)
```

#### ii. Max SWE DOY plot

```{r}
swe.maxDOY.comb <- ggplot(filter(snotelStats, metricName == "sweMax_doy"), aes(x = year, y = metricValue, color = elev))+
  geom_point()+
  geom_smooth(se = T, method = "lm", linetype = "dashed", color = "black")+
  theme_classic()+
  labs(x = "Water Year", y = "Day of year of maximum recorded SWE", color = "Site elevation, m")+
  scale_color_continuous(palette = elevPal)+
  geom_text(data = filter(cPVs, metricName == "sweMax_doy"), aes(x = x, label = paste("P=", label)), y = Inf, vjust = 2, color = "black")+
  theme(axis.text = element_text(size = 11), 
        axis.title = element_text(size = 11, face = "bold"), 
        strip.text = element_text(size = 10, face = "bold"), 
        legend.position = "bottom")


swe.maxDOY.comb

#Save plot:
ggsave("SWE_maxDOY_comb.jpg", swe.maxDOY.comb, device = "jpeg", path = here("outputs", "plots"), units = "in", dpi = "retina", height = 8, width = 11)
```

#### iii. Melt duration plot

```{r}
melt.comb <- ggplot(filter(snotelStats, metricName == "meltDays"), aes(x = year, y = metricValue, color = elev))+
  geom_point()+
  geom_smooth(se = T, method = "lm", linetype = "dashed", color = "black")+
  theme_classic()+
  labs(x = "Water Year", y = "# days between peak SWE and 1st day with 0 recorded SWE", color = "Site elevation, m")+
  scale_color_continuous(palette = elevPal)+
  geom_text(data = filter(cPVs, metricName == "meltDays"), aes(x = x, label = paste("P=", label)), y = Inf, vjust = 2, color = "black")+
  theme(axis.text = element_text(size = 11), 
        axis.title = element_text(size = 11, face = "bold"), 
        strip.text = element_text(size = 10, face = "bold"), 
        legend.position = "bottom")


melt.comb

#Save plot:
ggsave("melt_comb.jpg", melt.comb, device = "jpeg", path = here("outputs", "plots"), units = "in", dpi = "retina", height = 8, width = 11)
```

##4. Snow:total precip ratio

### A. Calculate snow:total precip ratio

Data prep & QA/QC - extract data with both air temp and precip; use empirical formula (Snow Probability = -50 * [tanh(0.4 * ( Mean Temp – 1.75))-1]) to calculate snow ratio: 

```{r}
rainSnow <- snotelClean%>%
  filter(!(is.na(temperature_mean)|is.na(precipitation)))%>% #drop rows without temp or precip data
  mutate(snowProb = round(-50*(tanh(0.4*(temperature_mean-1.75))-1), 3)/100
         )%>% #calculate snow probability using  method from Dai 2008, parameters from Tronstead et al. 2024. NOTE - had to divide by 100 to get numbers that made any sense. 
  mutate(dailySnow = precipitation*snowProb, #calculate daily snow amount as daily precip * snow probability
         dailyRain = precipitation*(1-snowProb))%>% #calculate daily rain amount as daily precip * (1-snow probability). Assumes no sleet. 
  mutate(precipCheck = dailySnow+dailyRain) #add column to check that rain amount + snow amount = daily precip amount. 
```

Next, calculate cumulative snow and cumulative total precip for each site/water Year

```{r}
snowStats <- rainSnow %>%
  group_by(site_name, elev, waterYear, siteElev)%>% #Group by site, water year, and elevation
  summarize(totPrecip = sum(precipitation), 
            totSnow = sum(dailySnow), 
            totRain = sum(dailyRain))%>% #calculate total annual precip, total annual snow, and total annual rain for each site/WY
  mutate(snowRatio = totSnow/totPrecip, #Calculate proportion of total precip falling as snow 
         totRainSnow = totSnow+totRain)%>% #Check that snow+rain = total annual precip
  filter(!is.nan(snowRatio)) #drop years w missing data
```

Separate calculation for December-Feb precip:

```{r}
winterSnow <- rainSnow %>%
  filter(month == 12|month == 1|month==2)%>%
  group_by(site_name, elev, waterYear, siteElev)%>% #Group by site, water year, and elevation
  summarize(totPrecip = sum(precipitation), 
            totSnow = sum(dailySnow), 
            totRain = sum(dailyRain))%>% #calculate total annual precip, total annual snow, and total annual rain for each site/WY
  mutate(snowRatio = totSnow/totPrecip, #Calculate proportion of total precip falling as snow 
         totRainSnow = totSnow+totRain)%>% #Check that snow+rain = total annual precip
  filter(!is.nan(snowRatio)) #drop years w missing data
```

### B. Run basic LM's and draw plots:

#### i. Entire water year:

```{r}
snowNest <- snowStats %>%
  nest(data = -c(site_name, elev, siteElev)) #create new nested dataframe with columns for site and metricName, one column for data - data entries for each site include year and metricValues. 
```

Run basic models for each site and metric of interest: 

```{r}
snowModels <- snowNest %>%
  mutate(model = map(data, ~lm(snowRatio~waterYear, data = .)%>% #run a model of temp~year for each site (referenced by "."); store results in new column called "model"
                     tidy))%>% #tidy function to construct a summary table with model results (seperate tables for each site)
  unnest(model) %>% #unnest model column to get rows for each site, columns for model metrics
  filter(term == "waterYear")  #remove intercept term to just get info on year for each site
snowModels
```

Add P-values for plotting: 

```{r}
snowPV <- merge(snowStats, 
                     select(snowModels, site_name, elev, siteElev, p.value)
                     )
saPVs <- snowPV%>%
  group_by(site_name, elev, siteElev)%>%
  summarize(x = mean(waterYear), label = round(unique(p.value), 2))
```

Annual snow proportion plot:

```{r}
annual.snow <- ggplot(snowStats, aes(x = waterYear, y = snowRatio, color = elev))+
  geom_point()+
  geom_smooth(se = T, method = "lm", linetype = "dashed")+
  theme_classic()+
  labs(x = "Water Year", y = "Snow:total water year precip ratio", color = "Elevation, m")+
  scale_color_continuous(palette = elevPal)+
  geom_text(data = saPVs, aes(x = x, label = paste("P=", label)), y = Inf, vjust = 2)+
  facet_wrap(~siteElev, scales = "free", labeller = label_wrap_gen(width = 20))+
  theme(axis.text = element_text(size = 11), 
        axis.title = element_text(size = 11, face = "bold"), 
        strip.text = element_text(size = 10, face = "bold"), 
        legend.position = "bottom")


annual.snow

#Save plot:
ggsave("waterYear_snowProp.jpg", annual.snow, device = "jpeg", path = here("outputs", "plots"), units = "in", dpi = "retina", height = 8, width = 11)
```

#### ii. December-February snow proportion

```{r}
winterNest <- winterSnow %>%
  nest(data = -c(site_name, elev, siteElev)) #create new nested dataframe with columns for site and metricName, one column for data - data entries for each site include year and metricValues. 
```

Run basic models for each site and metric of interest: 

```{r}
winterModels <- winterNest %>%
  mutate(model = map(data, ~lm(snowRatio~waterYear, data = .)%>% #run a model of temp~year for each site (referenced by "."); store results in new column called "model"
                     tidy))%>% #tidy function to construct a summary table with model results (seperate tables for each site)
  unnest(model) %>% #unnest model column to get rows for each site, columns for model metrics
  filter(term == "waterYear")  #remove intercept term to just get info on year for each site
winterModels
```

Add P-values for plotting: 

```{r}
winterPV <- merge(winterSnow, 
                     select(winterModels, site_name, elev, siteElev, p.value)
                     )
swPVs <- winterPV%>%
  group_by(site_name, elev, siteElev)%>%
  summarize(x = mean(waterYear), label = round(unique(p.value), 2))
```

Annual snow proportion plot:

```{r}
winter.snow <- ggplot(winterSnow, aes(x = waterYear, y = snowRatio, color = elev))+
  geom_point()+
  geom_smooth(se = T, method = "lm", linetype = "dashed")+
  theme_classic()+
  labs(x = "Water Year", y = "Dec - Feb Snow:total precip ratio", color = "Elevation, m")+
  scale_color_continuous(palette = elevPal)+
  geom_text(data = swPVs, aes(x = x, label = paste("P=", label)), y = Inf, vjust = 2)+
  facet_wrap(~siteElev, scales = "free", labeller = label_wrap_gen(width = 20))+
  theme(axis.text = element_text(size = 11), 
        axis.title = element_text(size = 11, face = "bold"), 
        strip.text = element_text(size = 10, face = "bold"), 
        legend.position = "bottom")


winter.snow

#Save plot:
ggsave("winter_snowProp.jpg", winter.snow, device = "jpeg", path = here("outputs", "plots"), units = "in", dpi = "retina", height = 8, width = 11)
```